The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Revisao Secao 3.

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in any conventional 
imperative language).

@@

We describe this relation with a set of \emph{small-step} structural semantics 
rules, which are built in such a way that at most one transition is possible at 
any time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(awaiting)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emitting)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}
%
An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
an artifact to remember the external sequence number $n$ it can awake.
An $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and its sequence number is smaller or 
equal than the current one ($m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emitting}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives the memory identifier and current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction chain, because 
a given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{awaiting}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   \\
\end{eqnarray*}
}
%
The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, defined in Figure~\ref{fig.isBlocked} and used in rules 
\textbf{and-adv2} and \textbf{or-adv2}, requiring the left branch $p$ to be 
blocked in order to allow the right transition from $q$ to $q'$.
%
Basically, the \emph{isBlocked} predicate determines that an expression becomes 
blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
expressions.

@@

For a parallel $and$, if one of the sides terminates, the composition is simply 
substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}).
The last two rules \textbf{and-brk1} and \textbf{and-brk2} deal with a $break$ 
in each of the sides in parallel.
A $break$ should terminate the whole composition in order to escape the 
innermost loop (\emph{aborting} the other side):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~mem(id)) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
\end{eqnarray*}
}
%
The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
all active $fin$ bodies of the side being aborted (to execute before the $and$ 
rejoins).
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
In Section~\ref{sec.formal.fins}, we show how a $fin$ is mapped to a 
finalization block in the concrete language.
%
Note that there is a syntactic restriction that a $fin$ body can only contain 
$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
reaction chain.

\begin{figure}
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( \_ )          &= mem(id)
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

For a parallel $or$, if one of the sides terminates, the whole composition 
terminates and applies the function $clear$ to the aborted side in order to 
properly finalize it (rules \textbf{or-nop1} and \textbf{or-nop2}).
Finally, a $break$ (rules \textbf{or-brk1} and \textbf{or-brk2}) behaves in the 
same way as in a parallel $and$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction chain.
%
However, $emitting$ expressions should resume their continuations of previous 
$emit$ in the ongoing reaction, they are just hanged in lower stack indexes 
(see rule \textbf{pop}).
%
Therefore, we define another relation that behaves as \textbf{relation-inner} 
(presented above) if the program is not blocked, and, otherwise, pops the 
stack:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(relation-outer)}
$$
%
To describe a \emph{reaction chain} in \CEU, i.e., how a program behaves in 
reaction to a single external event, we use the reflexive transitive closure of 
\textbf{relation-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we need multiple 
``invocations'' of reaction chains, incrementing the sequence number:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
& ...
\end{align*}
%
Each invocation starts with an external event at the top of the stack and 
finishes with a modified program and an empty stack.
After each invocation, the sequence number is incremented.

>>>>

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in conventional 
imperative languages).

@@

We describe this relation with a set of small-step structural semantics rules, 
which are built in such a way that at most one transition is possible at any 
time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}
%
An $await$ is simply transformed into an $awaiting$ (rule \textbf{await}) as an 
artifact to remember the external sequence number $n+1$ it can awake:
an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and it was reached in a previous 
reaction (i.e., sequence number $m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives a memory identifier and the current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction chain, because 
a given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   \\
\end{eqnarray*}
}
%
The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
\textbf{and-adv2} and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right transition from $q$ to $q'$.
%
Basically, the \emph{isBlocked} predicate determines that an expression becomes 
blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
expressions that cannot advance.

@@

For a parallel $and$, if one of the sides terminates, the composition is simply 
substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}).
The last two rules \textbf{and-brk1} and \textbf{and-brk2} deal with a $break$ 
in each of the sides in parallel.
To escape the innermost loop, a $break$ aborts the other side by applying the 
$clear$ function (to be described further):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~mem(id)) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
\end{eqnarray*}
}
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to the aborted side (rules 
\textbf{or-nop1} and \textbf{or-nop2}).
Finally, a $break$ (rules \textbf{or-brk1} and \textbf{or-brk2}) behaves in the 
same way as in a parallel $and$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
all active $fin$ bodies of the side being aborted, so that they execute before 
the composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
the concrete language to a $fin$ in the formal semantics.
%
Note that there is a syntactic restriction that a $fin$ body can only contain 
$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
reaction chain.

\begin{figure}
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( mem(id) )     &= mem(id)             \\
  clear( \_ )          &= nop
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction chain.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventualy resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another relation that behaves as \textbf{relation-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(relation-outer)}
$$
%
To describe a \emph{reaction chain} in \CEU, i.e., how a program behaves in 
reaction to a single external event, we use the reflexive transitive closure of 
\textbf{relation-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reaction chains in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with an external event at the top of the stack and 
finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.
